Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study

### 2015-05-14 chemotaxis/paper/draft1.tex
chemotaxis : yannis edits

\author[princeton]{Carmeline~J.~Dsilva}
\ead{cdsilva@princeton.edu}
%
\author[technion]{Ronen~Talmon}
\ead{ ronen@ee.technion.ac.il}
%
\author[yale]{Ronald R. Coifman}
\ead{coifman@math.yale.edu}
%
\author[princeton,princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
\ead{yannis@princeton.edu}

@@

\begin{abstract}
Nonlinear manifold learning algorithms have been effective for the analysis of complex data sets. 
%
However, these algorithms still encounter several challenges when faced with real data.
%
A key shortcoming is the existence of repeated eigendirections, where several embedding coordinates parameterize the same direction in the data set, therefore obscuring the true dimensionality of the underlying manifold. 
%
In this paper, we propose an algorithm, based on local linear regression, to automatically detect which coordinates are repeated eigendirections. 
%
We propose using only the eigenvectors which correspond to unique eigendirections to embed the data, and show that this reduced diffusion maps embedding induces a metric which is equivalent to the standard diffusion distance. 
%
We first demonstrate the utility and flexibility of our approach on synthetic data sets. 
%
We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis. 
%
Here, we show that our approach for factoring out repeated eigendirections can allow us to detect changes in dynamical behavior and the underlying system dimensionality directly from data. 
%
\end{abstract}

@@

In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{gepshtein2013image, fernandez2014diffusion, singer2011viewing, yuan2014automated, zhao2003face, trapnell2014dynamics, kemelmacher2011exploring, sifre2013rotation}. 
%
In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
%
For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%
For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies gives rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.

Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
%
Most manifold learning algorithms construct parameterizations of the data through the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
%
The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
%
Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale {\em ab initio} model.

These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}. 
%
Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally,rohrdanz2011determination.}.
%
A critical shortcoming of these methods is that from a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are ``repeated eigendirections'' which describe the same mode. 
%
Identifying those eigenvectors is critical for obtaining a parsimonious parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
%
This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
%
We first demonstrate our algorithm on synthetic examples where a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator is known, which will allow us to validate our proposed approach.
%
We then consider a complex data set from a stochastic dynamical system simulation which models cellular chemotaxis \cite{othmer1988models}. 
%
The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which successfully analyzes data from this simulation.
%
This pipeline allows us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.

\section{Manifold Learning Based on Laplace Operators}

Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
%
In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics.
%
Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$. 
%
We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\label{subfig:strip_efuncs}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\label{subfig:strip_evecs_uniform}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\label{subfig:strip_evecs_nonuniform}
\end{subfigure}
%
\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$. (b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian. (c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian. Note that in all cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.}
\end{figure}

We start by considering a continuous setting where rigorous analysis is available for specific examples. 
%
Using such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial. 
%
The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis.

We consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes, the principal axes parameterize the manifold.
%
However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent. 
%
In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
%
To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}). 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper. 
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
%
It was shown in \cite{coifman2006geometric} that, in the limit of infinite data, the discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the discrete Laplacian matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}

However, the data need not be sampled uniformly from the underlying manifold.
%
If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
$%\begin{equation}
(I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
$ %\end{equation}
where $U = -2 \log q$. 
%
Alternatively, one can factor out the density effects in the weight matrix.
%
Given data $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix
%
\begin{equation}
\tilde{W} = D^{-1} W D^{-1},
\end{equation}
%
where pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
%
We then construct the diagonal matrix $\tilde{D} \in \mathbb{R}^{m \times m}$, with $\tilde{D}_{ii} = \sum_j \tilde{W}_{ij}$, and the matrix $\tilde{A}  = \tilde{D}^{-1} \tilde{W}.$
%
The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator on $\mathcal{M}_d$ even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.

As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows data sampled from a strip, colored by eigenvectors of $A$. 
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.  
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these two eigenvectors are guaranteed to appear before any coupled or repeated eigendirections. 
%
However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, and identifying which eigenvectors correspond to unique eigendirections is nontrivial.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections. 
%
This problem was addressed previously in \cite{gerber2007robust} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
%
However, this algorithm is {\em ad hoc}, and both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
%TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc). 
%
Here, we propose an alternative approach to address the problem of repeated eigendirections. 
%
Our approach is motivated by simple trigonometric arguments, such as the fact that $\cos2x$ (a repeated eigendirection) can be written as a function of $\cos x$ (a unique eigendirection). 
%
We therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$; if the resulting fit is accurate, we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
%
We use a local linear function 
\begin{equation}
\phi_k \approx \alpha + \beta^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.

>>>>

\author[princeton]{Carmeline~J.~Dsilva}
%\ead{cdsilva@princeton.edu}
%
\author[technion]{Ronen~Talmon}
%\ead{ ronen@ee.technion.ac.il}
%
\author[yale]{Ronald R. Coifman}
%\ead{coifman@math.yale.edu}
%
\author[princeton,princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
\ead{yannis@princeton.edu}

@@

\begin{abstract}
%
Nonlinear manifold learning algorithms, such as diffusion maps, have been fruitfully
applied in recent years to the analysis of large and complex data sets.
%
However, such algorithms still encounter challenges when faced with real data.
%
One such 
challenge is the existence of ``repeated eigendirections," which obscures the detection of the 
true dimensionality of the underlying manifold and arises when several embedding coordinates parametrize
the same direction in the intrinsic geometry of the data set. 
%
We propose an algorithm, based on local linear regression, 
to automatically detect coordinates corresponding to repeated eigendirections.
%
We construct a more parsimonious embedding using only the eigenvectors corresponding to unique eigendirections, and we show that this reduced diffusion maps embedding induces 
a metric which is equivalent to the standard diffusion distance.
%
We first demonstrate the utility and flexibility of our approach on synthetic data sets.
%
We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
intrinsic system dimensionality directly from data.
%
\end{abstract}

@@

In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{gepshtein2013image, fernandez2014diffusion, singer2011viewing, yuan2014automated, zhao2003face, trapnell2014dynamics, kemelmacher2011exploring, sifre2013rotation}.
%
For multiscale dynamical systems in particular, data-driven methodologies are essential for
assisting in model reduction when simple macroscopic models cannot be analytically obtained
due to the system complexity \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
%
For simulations or experimental observations of such systems, the detailed, microscale evolving system state is
very high-dimensional, and the reduction to useful macroscale dynamics is not obvious.
%
In such cases, data obtained from observations and/or simulations of the dynamical system combined with
data mining methodologies can lead to a low-dimensional description which not only provides insight into the underlying dynamics, but also serves as a first step in constructing macroscale models consistent with the observed microscale behavior.

%Due to the complexity and range of microscale behaviors possible in dynamical systems,
%we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
%
Several manifold learning algorithms obtain parametrizations of the data through
the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
%
The data is then embedded in a new low-dimensional coordinate system
given by the eigenvectors of this operator.
%
The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables",
the variables which parametrize the macroscale dynamical behavior of the system, thus enabling the construction of a reduced macroscale model.

Such manifold learning algorithms were initially applied to synthetic data sets, to illustrate
their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}.
%
More recently they have been applied to experimental and simulation data,
enabled by advances in data representation (observers) and
metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination}.
%
A nontrivial shortcoming of these methods is that, from a geometric perspective,
not all eigenvectors are guaranteed to parametrize unique directions within the data;
some eigenvectors are ``repeated eigendirections'' which describe the same coordinate in the intrinsic geometry
of the data.
%
Identifying these eigenvectors is critical for obtaining a
parametrization of the system which captures the true dimensionality of the macroscale dynamics.
%
This task is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
%
We first demonstrate and validate our algorithm on synthetic examples, where a closed-form solution for the
eigenvectors and eigenvalues of the Laplace operator is known.
%
We then consider a simulation data set from a stochastic dynamical system modeling
cellular chemotaxis \cite{othmer1988models}.
%
Recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections,
provide a data analysis pipeline which successfully analyzes this simulation in a purely data-driven manner.
%
We will show that this pipeline allows us to detect changes in the dimensionality of the underlying macroscale dynamics, which are
related to changes in the regime/mode of the system.

\section{Manifold Learning Based on Laplace Operators}
%Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description.
%
%In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parametrization of this manifold corresponds to parametrizing the low-dimensional, macroscale dynamics.
%

Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ state observations sampled from an evolving autonomous dynamical system.
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$.
%
We therefore consider the problem of parametrizing the continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$ from data.
%
For linear hyperplanes, the principal axes parametrize the manifold.
%
However, in the case when the manifold is nonlinear, a set of appropriate coordinates is not readily apparent.
%
By using a particular manifold learning technique based on the construction of a Laplace operator, namely, diffusion maps, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_cnts1}
\includegraphics[width=\textwidth]{strip_cnts2}
\caption{}
\label{subfig:strip_efuncs}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_discrete1}
\includegraphics[width=\textwidth]{strip_discrete4}
\caption{}
\label{subfig:strip_evecs_uniform}
\end{subfigure}
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{strip_nonuniform1}
\includegraphics[width=\textwidth]{strip_nonuniform2}
\caption{}
\label{subfig:strip_evecs_nonuniform}
\end{subfigure}
%
\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$. (b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian. (c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian. Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$.}
\end{figure}

In recent years it has been noted that the eigenfunctions of the \emph{continuous}
Laplace-Beltrami operator provide ``good" coordinates for a manifold \cite{jones2008}.
%
We start by considering such a continuous setting, where rigorous analysis is possible for specific examples.
%
Using such an example, we will demonstrate that the existence of repeated eigendirections is
inherent to the manifold learning setup based on Laplace operators,
and that the identification of the unique eigendirections is nontrivial.
%
The fact that these challenges arise in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for data analysis.

To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}.
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
respectively, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
(this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper.
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
smallest $d$ eigenvalues.
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
Clearly, this problem of repeated eigendirections also arises for more complex manifolds of more than two variables.
%
We note that, although the eigenfunctions can only be written analytically for very special cases,
it has been observed empirically that they often provide appropriate coordinates
to usefully parametrize more complex, nonlinear manifolds.
%

@@

In most applications, we are not given a description of the continuous manifold on which the data lie.
%
Instead, we are given data {\em sampled} from the underlying manifold,
and the parameterization of the manifold needs to be uncovered from the data.
%
This can be accomplished by constructing a matrix which approximates the Laplace-Beltrami operator.
%
It was shown in \cite{coifman2006geometric} that, in the limit of infinite data,
this discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $\mathbf{W} \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
\mathbf{W}_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations.
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{rohrdanz2011determination, coifman2008graph};
we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{m \times m}$, with $\mathbf{D}_{ii} = \sum_j \mathbf{W}_{ij}$,
and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 < \alpha < 1$.
%
Next, we construct the diagonal matrix $\widetilde{\mathbf{D}} \in \mathbb{R}^{m \times m}$, with $\widetilde{\mathbf{D}}_{ii} = \sum_j \widetilde{\mathbf{W}}_{ij}$, and the matrix $\mathbf{A}  = \widetilde{\mathbf{D}}^{-1} \widetilde{\mathbf{W}}.$

If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then, for $\epsilon \rightarrow 0$ and $m \rightarrow \infty$ (with the appropriate rates), the discrete matrix converges to the following continuous limit operators with Neumann boundary conditions (as discussed in the previous section) \cite{coifman2006geometric}
\begin{align}
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - 2\nabla U \cdot \nabla \phi, &&\alpha = 0 \\
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, &&\alpha = 1/2 \\
\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi, &&\alpha = 1
\end{align}
where $U = - \log q$.
%
The different limit operators, depending on the choice of $\alpha$, imply that nonuniform
sampling on the manifold may have different effects.
%
In particular, by setting $\alpha=1$, one can factor out the density effects in the
weight matrix, and the discrete Laplacian matrix approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$.
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $\mathbf{A}$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $\mathbf{A}$ are related to the eigenvalues of the continuous operator by
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}

As discussed previously, the eigenfunctions provide a parametrization of the manifold,
such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \ge 0$ is
a parameter (in our examples, we will take $\tau=0$).
%
This is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
we order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
%
Because the matrix $\mathbf{A}$ is row-stochastic ($\sum_j \mathbf{A}_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
The distance induced by this mapping is called the standard diffusion distance,
%
\begin{equation}
D^2_\tau(z(i), z(j)) = \sum_{k=1}^{m-1} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2.
\end{equation}
%
Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows
data sampled from a strip, colored by eigenvectors of $\mathbf{A}$.
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$,
and thus parametrize the manifold.
%
%TODO: maybe we can put here the new figures for the different values of $\alpha$ instead of these figures.
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$
provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
%
From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of
the corresponding eigenvalues implies that these two eigenvectors are guaranteed to
appear before any coupled or repeated eigendirections.
%
However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors,
as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, so that identifying which eigenvectors correspond to unique eigendirections is nontrivial.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce
which ones capture new directions in the data, and which ones are merely repeated eigendirections.
%
This problem was addressed previously in \cite{gerber2007robust} by performing
successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration.
%
However, both the advection
procedure and the successive eigendecompositions are expensive and soon become intractable for larger data sets, 
and the approach is somewhat {\em ad hoc}.
%TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc).
%
Here, we propose an alternative approach to address the problem of repeated eigendirections
motivated by simple trigonometric arguments (for example, the repeated eigendirection $\cos2x$ 
can be written as a function of the unique eigendirection $\cos x$).
%
We therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$; if the
resulting fit is accurate, we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$.
%
We use a {\em local} linear function
\begin{equation}
\phi_k \approx \alpha + \beta^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$.
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.
